---
title: "Reproducible Research - Course Project 1"
author: "Alexander Baker"
date: '2017-12-05'
output: html_document
---

## Loading and Preprocessing the Data

The first step is to clear the data environment, download the data, load it into R and process it into a format suitable for analysis.

```{r}
activity <- read.csv(file="activity.csv", head=TRUE,sep=",")
summary(activity)
```

```{r}
library(ggplot2) #load ggplot2 for later plotting
library(lattice) #load lattice for later plotting
```

## What is the mean and median total number of steps taken per day?

Next we will determine (a) the total number of steps taken per day, prepare (b) a histogram of the total number of steps taken each day and calculate (c) the mean and median total number of steps taken per day.

```{r}
#the total number of steps taken per day
stepsperday <- aggregate(steps ~ date, activity, sum)
head(stepsperday)
```


```{r}
# a histogram of the total number of steps taken each day
hist(stepsperday$steps, main = paste("Histogram of Frequency and Total Steps per Day"), col="red", xlab="Total Steps per Day")
```


```{r}
# the mean and median total number of steps taken per day
perdmean <- mean(stepsperday$steps)
perdmean

```{r}
perdmedian <- median(stepsperday$steps)
perdmedian
```

## What is the average daily activity pattern?

We can see from a time series plot the 5-minute interval and the average number of steps taken, averaged across all days.

```{r}
avg_steps <- aggregate(activity$steps, by=list(activity$interval), mean, na.rm=TRUE)
names(avg_steps) <- c("interval", "mean")
plot(avg_steps$interval, avg_steps$mean, type = "l", col="green", lwd = 2, xlab="Interval", ylab="Average Number of Steps", main="Average Number of Steps per Interval")
```



The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is:
```{r}
avg_steps[which.max(avg_steps$mean), ]$interval
```
835.

## Imputting Missing Values

There are a number of days and intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into calculations.

In response, we will (a) determine the total number of missing values in the dataset, (b) execute a strategy for filling all of the missing values in the dataset by 
```{r}
# (a) determine the total number of missing values in the dataset
sum(is.na(activity))
```

```{r}
# (b) execute a strategy for filling all of the missing values in the dataset and create a new dataset equal to the original dataset with the missing data filled in.

imputed_pre <- avg_steps$mean[match(activity$interval, avg_steps$interval)]
imputed_act <- transform(activity, steps = ifelse(is.na(activity$steps), yes = imputed_pre, no = activity$steps))
imputed_total <- aggregate(steps ~ date, imputed_act, sum)
names(imputed_total) <- c("date", "steps")
```

``` {r}
hist(imputed_total$steps, col = "red", xlab = "Steps per Day", ylim = c(0,30), main = "Imputed Histogram of Frequency and Total Steps per Day", breaks = seq(0,25000,by=2500))
```

The imputed mean and median are:

```{r}
imputed_mean <- mean(imputed_total$steps)
imputed_mean
```
```{r}
imputed_median<- median(imputed_total$steps)
imputed_median
```

Both the mean and median have risen to reflect the addition of missing values, albeit the mean rising more than the median. These imputed means and medians are still within a few hundred of the original values.

## Are there differences in activity patterns between weekdays and weekends?

In order to discover whether or not there are differences between weekday and weekend, we will create a new factor variable and make a panel plot of the average number of steps taken averaged across weekday and weekend days.

```{r}
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))
activity$datetype <- sapply(activity$date, function(x) {
        if (weekdays(x) == "Saturday" | weekdays(x) =="Sunday") 
                {y <- "Weekend"} else 
                {y <- "Weekday"}
                y
        })
head(activity)
```

```{r}
activity_by_date <- aggregate(steps~interval + datetype, activity, mean, na.rm = TRUE)
plot<- ggplot(activity_by_date, aes(x = interval , y = steps, color = datetype)) +
       geom_line() +
       labs(title = "Average Daily Steps by Weekday or Weekend", x = "Interval", y = "Average Number of Steps") +
       facet_wrap(~datetype, ncol = 1, nrow=2)
print(plot)
```

We can see from this data that there is an early and higher peak during weekdays, but higher sustained activity on weekends.